Linear stability of magnetized massive protoplanetary disks

### 2014-06-03 paper/cover.tex

I am writing to submit the manuscript `Linear stability of magnetized massive protoplanetary disks'.

>>>>

I am writing to submit the revised manuscript `Linear stability of
magnetized massive protoplanetary disks', in response to the recent
reviewer's report. Revisions are made in bold, and I attach my
response to the reviewer.

### 2014-06-03 paper/results.tex
final revised version

\subsubsection{Uniform
  resistivity}   
Fig. \ref{compare_growth_poly_uniresis} plots MRI growth rates as a
function of $Q$ and $\Lambda_0$. The resistivity is uniform ($A=1$). 
For ideal MHD and {\bf a weak field ($\Lambda_0>1$, $\beta=100$)}, there is
negligible dependence on $Q$. {\bf However, with $\beta=25$ or 
in the resistive limit  ($\Lambda_0<1$)}, growth rates decrease noticeably for $Q<0.5$
($Q_\mathrm{2D}\lesssim 1.5$).  Since we find density and potential
perturbations to be negligible (i.e. the linear response is
non-self-gravitating), this shows that disk self-gravity can affect the
MRI through the background equilibrium.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_growth_poly_uniresis2}
  \caption{{\bf MRI growth rates as a function of $Q$ and midplane 
    Elsasser numbers $\Lambda_0$, in polytropic disks with $\beta=100$
    (solid, dotted and dot-dashed lines). The dash-triple dot line is
    the $\Lambda_0=10$ case with $\beta=25$. The resistivity is
    uniform.}   
    \label{compare_growth_poly_uniresis}}
\end{figure}

@@

In the ideal limit {\bf with $\beta=100$}, we find $\lambda < 2H$
throughout most of the disk for the values of $Q$ considered, so
self-gravity does not affect growth rates significantly. However, the
ratio $\lambda/2H$ does increase with stronger
self-gravity. Consequently, the wavelength of the instability, in
units of $H$, increases. This is shown in
Fig. \ref{compare_result_lambda10} which plots the magnetic energies
for $\Lambda_0=10$ and a range of $Q$ values. The number of vertical 
nodes decrease with $Q$, i.e. the disk accommodates fewer wavelengths
because increasing vertical self-gravity makes it thinner.

{\bf
We repeated the $\Lambda_0=10$ case with a stronger field $\beta=25$,
shown in Fig. \ref{compare_growth_poly_uniresis} as the dashed-triple
dot line. Here, strong self-gravity is effective in reducing the
growth rate, because decreasing $\beta$ enhances the dependence of
$\lambda/H$ on $f(Q)$. 
}

@@

Self-gravity {\bf also} appreciably decreases the MRI growth rates in the
resistive limit. Fig. \ref{lambda_poly_resis} plots 
Eq. \ref{sano_crit} for $\Lambda_0=0.3$. In the non-self-gravitating
disk ($Q=4$) the instability criterion is marginally satisfied and the
MRI operates. As $Q$ decreases, 
Eq. \ref{sano_crit} is violated and the MRI growth rate is
significantly reduced. This is seen for $Q=0.2$ where $\lambda \geq 2H$ throughout
the disk. (The instability is not suppressed since
Eq. \ref{lambda_ideal}---\ref{lambda_resis} is only exact for
unstratified disks.) Although the function 
$f(Q)$ does not change significantly for the range of $Q$ considered,
the dependence of {\bf $\lambda/H$} on $f(Q)$ is amplified by the
denominator $\Lambda_0<1$ in the resistive case. Modes in
Fig. \ref{lambda_poly_resis} have no nodes in the magnetic energy
$E_m$ except at $z\simeq0,\,H$, i.e. only the longest wavelength survives
{\bf against} large resistivity. 
%self-gravity tips it over

@@

{\bf 
%We find qualitatively similar results for large $A$ with small
%  $\Lambda_0$. 
We also performed additional calculations with $A=10^3$ and $A=10^4$
\citep[see Fig. 1 of ][]{gressel12}. For $\Lambda_0=0.1$, we find no
significant increase in the magnetic energy in the resistive zones. 
However, lowering $\Lambda_0$ gives similar
results to Fig. \ref{poly_layer}, e.g. for $A=10^3$ and
$\Lambda_0=10^{-2}$ or $A=10^4$ and $\Lambda_0=10^{-3}$ the magnetic
energy penetrates into the resistive zone for strongly
self-gravitating disks. In general, the magnetic energy density maximum moves
toward the midplane with increasing self-gravity.    
}

>>>>

\subsubsection{Uniform
  resistivity}   
Fig. \ref{compare_growth_poly_uniresis} plots MRI growth rates as a
function of $Q$ and $\Lambda_0$. The resistivity is uniform ($A=1$). 
For ideal MHD and a weak field ($\Lambda_0>1$, $\beta=100$), there is
negligible dependence on $Q$. However, with $\beta=25$ or 
in the resistive limit  ($\Lambda_0<1$), growth rates decrease noticeably for $Q<0.5$
($Q_\mathrm{2D}\lesssim 1.5$).  Since we find density and potential
perturbations to be negligible (i.e. the linear response is
non-self-gravitating), this shows that disk self-gravity can affect the
MRI through the background equilibrium.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_growth_poly_uniresis2}
  \caption{ MRI growth rates as a function of $Q$ and midplane 
    Elsasser numbers $\Lambda_0$, in polytropic disks with $\beta=100$
    (solid, dotted and dot-dashed lines). The dash-triple dot line is
    the $\Lambda_0=10$ case with $\beta=25$. The resistivity is
    uniform.   
    \label{compare_growth_poly_uniresis}}
\end{figure}

@@

In the ideal limit with $\beta=100$, we find $\lambda < 2H$
throughout most of the disk for the values of $Q$ considered, so
self-gravity does not affect growth rates significantly. However, the
ratio $\lambda/2H$ does increase with stronger
self-gravity. Consequently, the wavelength of the instability, in
units of $H$, increases. This is shown in
Fig. \ref{compare_result_lambda10} which plots the magnetic energies
for $\Lambda_0=10$ and a range of $Q$ values. The number of vertical 
nodes decrease with $Q$, i.e. the disk accommodates fewer wavelengths
because increasing vertical self-gravity makes it thinner.

We repeated the $\Lambda_0=10$ case with a stronger field $\beta=25$,
shown in Fig. \ref{compare_growth_poly_uniresis} as the dashed-triple
dot line. Here, strong self-gravity is effective in reducing the
growth rate, because decreasing $\beta$ enhances the dependence of
$\lambda/H$ on $f(Q)$. For $Q=0.2$ and $\beta=25$ we find 
$\lambda_\mathrm{ideal}/2H\sim 1$ at the midplane and the growth rate is reduced significantly.

@@

Self-gravity also appreciably decreases the MRI growth rates in the
resistive limit. Fig. \ref{lambda_poly_resis} plots 
Eq. \ref{sano_crit} for $\Lambda_0=0.3$. In the non-self-gravitating
disk ($Q=4$) the instability criterion is marginally satisfied and the
MRI operates. As $Q$ decreases, 
Eq. \ref{sano_crit} is violated and the MRI growth rate is
significantly reduced. This is seen for $Q=0.2$ where $\lambda \geq 2H$ throughout
the disk. (The instability is not suppressed since
Eq. \ref{lambda_ideal}---\ref{lambda_resis} is only exact for
unstratified disks.) Although the function 
$f(Q)$ does not change significantly for the range of $Q$ considered,
the dependence of $\lambda/H$ on $f(Q)$ is amplified by the
denominator $\Lambda_0<1$ in the resistive case. Modes in
Fig. \ref{lambda_poly_resis} have no nodes in the magnetic energy
$E_m$ except at $z\simeq0,\,H$, i.e. only the longest wavelength survives
 against large resistivity. 
%self-gravity tips it over

@@

%We find qualitatively similar results for large $A$ with small
%  $\Lambda_0$. 
We also performed additional calculations with $A=10^3$ and $A=10^4$
\citep[see Fig. 1 of ][]{gressel12}. For $\Lambda_0=0.1$, we find no
significant increase in the magnetic energy in the resistive zones. 
However, lowering $\Lambda_0$ gives similar
results to Fig. \ref{poly_layer}, e.g. for $A=10^3$ and
$\Lambda_0=10^{-2}$ or $A=10^4$ and $\Lambda_0=10^{-3}$ the magnetic
energy penetrates into the resistive zone for strongly
self-gravitating disks. In general, the magnetic energy density maximum moves
toward the midplane with increasing self-gravity.

### 2014-06-03 paper/results_azi.tex

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0cm
    0cm]{figures/compare_growth3_tilted_even.ps}  
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0.52cm]{figures/compare_growth3_tilted_odd.ps} 
  \caption{MRI growth rates in isothermal disks with $Q=~0.2$ ($Q_\mathrm{2D}=0.72$) and 
    $\beta=10$ for a range of azimuthal field strengths $B_y/B_z$. The
    dots are solutions computed from the full problem, with the
    colorbar measuring the gravitational potential perturbation {\bf
      via $\tau$}, while
    the solid curves are computed from the Cowling approximation. 
    For $B_y=0$, modes in top and bottom panels have
    $W^\prime(0)=0$ and $W(0)=0$, respectively.
    \label{compare_growth3_tilted}}
\end{figure}

@@

The modes in the bottom panel of Fig. \ref{compare_growth3_tilted} 
display opposite behavior. For $B_y=0$ we find {\bf$W(z)$} is odd, and
self-gravity has no effect. When $B_y>0$, $W$ deviates from an
odd function and the midplane density perturbation $|W(0)|$
increases. SG is stabilizing for these modes at all wavelengths, and
is most effective at $k_xH =  O(1)$. 
Fig. \ref{result_tilted} show eigenfunctions for $\epsilon=3$ and
$k_xH=1.1$ with and without the Cowling approximation.
SG significantly enhances the midplane density perturbation, making the
gravitational potential energy comparable to the magnetic energy, 
which becomes more confined near the midplane.

>>>>

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0cm
    0cm]{figures/compare_growth3_tilted_even.ps}  
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0.52cm]{figures/compare_growth3_tilted_odd.ps} 
  \caption{MRI growth rates in isothermal disks with $Q=~0.2$ ($Q_\mathrm{2D}=0.72$) and 
    $\beta=10$ for a range of azimuthal field strengths $B_y/B_z$. The
    dots are solutions computed from the full problem, with the
    colorbar measuring the gravitational potential perturbation 
      via $\tau$, while
    the solid curves are computed from the Cowling approximation. 
    For $B_y=0$, modes in top and bottom panels have
    $W^\prime(0)=0$ and $W(0)=0$, respectively.
    \label{compare_growth3_tilted}}
\end{figure}

@@

The modes in the bottom panel of Fig. \ref{compare_growth3_tilted} 
display opposite behavior. For $B_y=0$ we find $W(z)$ is odd, and
self-gravity has no effect. When $B_y>0$, $W$ deviates from an
odd function and the midplane density perturbation $|W(0)|$
increases. SG is stabilizing for these modes at all wavelengths, and
is most effective at $k_xH =  O(1)$. 
Fig. \ref{result_tilted} show eigenfunctions for $\epsilon=3$ and
$k_xH=1.1$ with and without the Cowling approximation.
SG significantly enhances the midplane density perturbation, making the
gravitational potential energy comparable to the magnetic energy, 
which becomes more confined near the midplane.
